Case for Support

### 2021-11-17 application/backgroundToTheResource.tex
Updates from Overleaf
Updates from Goncalo

  \caption{\footnotesize Bonsai development timeline and community adoption. (A) Timeline of
    milestones and landmark package releases. (B) Example capabilities 
    within Bonsai. (C) Bonsai workflow for tracking the position of objects
    in a video. (D) Worldwide
    distribution of visitors to the bonsai-rx.org website. (E)
    Number of downloads for each Bonsai version. (F) Cumulative number of
    registered forum users, citations of the main publication, and number of
    training events, by year (bottom panels), versus release and development
    cycle (top panel), measured in commits to main repository. 
    % Click on the image to see its online version.
    }

@@

To understand the neural basis of behaviour, experiments need to control and quantify animal behaviour whilst recording, and often manipulating, neural activity. 
%
Powerful experimental designs require behavioural feedback (e.g.\ reward, lights, sounds, actuated
devices) or neural perturbation (e.g.\ using light-sensitive polarising
molecules) that depend on real-time behaviour or neural state.
%
Furthermore, it has become increasingly apparent that the neural signals that
drive complex behaviour are manifest in a broad population of neurons rather
than in single cells. Current neural recording emphasises high-dimensional
signals from 100s to 1000s of neurons monitored using multi-electrode arrays or cellular-resolution imaging, with a wide array of probe configurations and neural targets.
%
\textbf{This complexity of experiments limits the potential for database-driven progress in systems neuroscience}. The number of possible experiments is astronomical, and new questions are typically answered by designing and conducting new experiments rather than mining the data from an old one.

At present, progress in the field is held back by the challenge of rapidly designing and implementing new experimental protocols, and the difficulty of integrating state-of-the-art data processing and
neuroinformatics into custom experimental designs.
%
\textbf{Reproducibility}, in particular, depends on the availability of standardised tools for experimental control and analysis, shared between laboratories (Baker, 2016; Ioannidis, 2005). Although such tools exist in genetics, astronomy, physics and medicine (Fish et al., 2016; Abdalla et al., 2018, CERN Education, Communications and Outreach Group, 2018, Dickinson et al., 2016, Bycroft et al., 2018), they are mostly lacking in neuroscience, with documented consequences for the reliability of ``discoveries'' (Baker, 2016; Botvinik-Nezer et al., 2020; Button et al., 2013).

@@

Bonsai achieves robust behavioural control, hardware interaction and signal acquisition while ensuring remarkable ease of use.  The proposed project will provide valuable enhancements.  We will develop, test and deploy a new software workflow making it possible to incorporate external packages into Bonsai's C\#-based environment.  Building on this new extensibility, we will then incorporate a suite of machine-learning-driven analysis tools into the Bonsai environment.  These added capabilities will allow  video-based behavioural analysis as well as real-time interfacing of population-scale activity with laboratory control.  In parallel we will develop the infrastructure needed to evaluate competing data analytic approaches within a single experimental framework

>>>>

  \caption{Bonsai development timeline and community adoption. (A) Timeline of
    milestones and landmark package releases. (B) Example capabilities 
    within Bonsai. (C) Bonsai workflow for tracking the position of objects
    in a video. (D) Worldwide
    distribution of visitors to the bonsai-rx.org website. (E)
    Number of downloads for each Bonsai version. (F) Cumulative number of
    registered forum users, citations of the main publication, and number of
    training events, by year (bottom panels), versus release and development
    cycle (top panel), measured in commits to main repository. 
    % Click on the image to see its online version.
    }

@@

To understand the neural basis of behaviour, experiments need to control and quantify animal behaviour whilst recording, and often manipulating, neural activity. 
%
Powerful experimental designs require behavioural feedback (e.g.\ reward, lights, sounds, actuated
devices) or neural perturbation (e.g.\ using light-sensitive polarising
molecules) that depend on real-time behaviour or neural state.
%
It has also become increasingly apparent that the neural signals that
drive complex behaviour are manifest in a broad population of neurons rather
than in single cells. Current neural recording emphasises high-dimensional
signals from 100s to 1000s of neurons monitored using multi-electrode arrays or cellular-resolution imaging, with a wide array of probe configurations and neural targets.
%
\textbf{This complexity of experiments limits the potential for database-driven progress in systems neuroscience}. The number of possible experiments is astronomical, and new questions are typically answered by designing and conducting new experiments rather than mining data from old ones.

At present, progress in the field is held back by the challenge of designing and implementing new experimental protocols, and the difficulty of integrating state-of-the-art data processing and neuroinformatics into custom experimental designs.
%
\textbf{Reproducibility}, in particular, depends on the availability of standardised tools for experimental control and analysis, shared between laboratories (Baker, 2016; Ioannidis, 2005). Although such tools exist in genetics, astronomy, physics and medicine (Fish et al., 2016; Abdalla et al., 2018, CERN Education, Communications and Outreach Group, 2018, Dickinson et al., 2016, Bycroft et al., 2018), they are mostly lacking in neuroscience, with documented consequences for the reliability of ``discoveries'' (Baker, 2016; Botvinik-Nezer et al., 2020; Button et al., 2013).

@@

Bonsai achieves robust behavioural control, hardware interaction and signal acquisition while ensuring remarkable ease of use.  The proposed project will provide valuable enhancements to the existing codebase.  We will develop, test and deploy a new software workflow making it possible to incorporate external packages into Bonsai's C\#-based environment.  Building on this new extensibility, we will then incorporate a suite of machine-learning-driven analysis tools into the Bonsai environment.  These added capabilities will allow  video-based behavioural analysis as well as real-time interfacing of population-scale activity with laboratory control.  In parallel we will develop the infrastructure needed to evaluate competing data analytic approaches within a single experimental framework

### 2021-11-17 application/bonsai.tex
Updates from Goncalo

Standard software tools for data analysis (ImageJ, MATLAB, R, etc.) have been
transformative to the progress of increasingly “data-rich” sciences. However,
the equivalent standardized software tools for data acquisition and control of
animal behaviour neuroscience experiments are still lacking. Life science experiments demand
a combination of multiple instrumentation and control technologies, for both
behavioural and physiological investigations. The growing complexity on both the
amount of data that is collected, and the rich conditions under which behaviour
must be explored, place an increasing burden on experimenters to integrate
highly specialised equipment in unique configurations, while often lacking
expertise in the relevant engineering fields.

Reproducibility is a key element of Bonsai's design.  Data acquisition and experimental control protocols can be 
replicated in any laboratory by just sharing a Bonsai configuration file.

@@

The language has also helped to potentiate the growing wave of foundational
open hardware initiatives, such as the OpenEphys \citep{siegleEtAl17} and UCLA
Miniscope~\citep{caiEtAl16}, making it possible to quickly combine and
integrate these tools into new experiments~\citep{buccinoEtAl18}.
%
Bonsai has been adopted in large neuroscience undertakings like the
International Brain
Laboratory\footnote{https://www.internationalbrainlab.com/}
and the Allen Institute for Brain
Science\footnote{https://alleninstitute.org/what-we-do/brain-science/}.
%
% LEAVE OUT FOR THIS GRANT:
% More recently, Bonsai has started to expand outside the domain of neuroscience
% into biomedical research and biotechnology tool development, and even outside
% academia into public outreach and education programs.

>>>>

% I think we've said something similar in the first para  - m.
%
% Standard software tools for data analysis (ImageJ, MATLAB, R, etc.) have
% transformed progress inf increasingly “data-rich” sciences. However,
% equivalent standardised software tools for data acquisition and control of
% animal behaviour in neuroscience experiments are still lacking. Life science experiments demand
% a combination of multiple instrumentation and control technologies, for both
% behavioural and physiological investigations. The growing complexity on both the
% amount of data that is collected, and the rich conditions under which behaviour
% must be explored, place an increasing burden on experimenters to integrate
% highly specialised equipment in unique configurations, while often lacking
% expertise in the relevant engineering fields.

@@

The language has also helped to potentiate the growing wave of foundational
open hardware initiatives, such as the OpenEphys \citep{siegleEtAl17} and UCLA
Miniscope~\citep{caiEtAl16}, making it possible to quickly combine and
integrate these tools into new experiments~\citep{buccinoEtAl18}.
%
Bonsai has been adopted in large neuroscience undertakings like the
International Brain
Laboratory\footnote{https://www.internationalbrainlab.com/}
and the Allen Institute for Brain
Science\footnote{https://alleninstitute.org/what-we-do/brain-science/}.
%
Key to the wide adoption is the reproducibility that Bonsai ensures.  Data acquisition and experimental control protocols can be 
replicated in any laboratory by just sharing a Bonsai configuration file. 
%
% LEAVE OUT FOR THIS GRANT:
% More recently, Bonsai has started to expand outside the domain of neuroscience
% into biomedical research and biotechnology tool development, and even outside
% academia into public outreach and education programs.

### 2021-11-17 application/caseForSupport.tex
Removed live links
Updates from Overleaf
Updates from Goncalo

\usepackage[colorlinks=true]{hyperref}
\usepackage{natbib}
\usepackage{apalike}
\usepackage{graphicx}
\usepackage{verbatim}

@@

@@

\section*{\large Case for support}
\vskip 0.5ex
We propose to extend, enhance, maintain and support \textbf{Bonsai}, a fully
integrated software environment to \textbf{enable cutting-edge reproducible
  laboratory systems neuroscience experiments in animal models}, with a
particular emphasis on \textbf{machine-intelligence-enabled, real-time
  neuroinformatics} methods.

>>>>

% \usepackage[colorlinks=true]{hyperref}
\usepackage[colorlinks = true,
            linkcolor = blue!50!black!50,
            urlcolor  = blue!50!black!50,
            citecolor = blue!50!black!50,
            anchorcolor = blue!50!black!50]{hyperref}
\usepackage{natbib}
\usepackage{apalike}
\usepackage{graphicx}
\usepackage{verbatim}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{calc}

@@

\usepackage[font={footnotesize}]{caption}

@@

\section*{\large Case for support}
\vskip 0.5ex
We propose to extend, enhance, maintain and support \textbf{Bonsai}, an integrated software environment to enable \textbf{cutting-edge reproducible
  laboratory systems neuroscience experiments in animal models}, with a
particular emphasis on \textbf{machine-intelligence-enabled, real-time
  neuroinformatics} methods.

### 2021-11-17 application/detailsOfResource.tex
Updates from Goncalo

@@

To avoid the need to re-implement a single data analysis algorithm in multiple different contexts we will create a set of abstract machine-intelligence operators in Bonsai that can be configured to map between data substrates and machine-learning algorithms.
%
The planned abstraction layer is based on the successful \texttt{scikit-learn} Python
library \citep{buitinckEtAl13}.
%
By analogy to the interfaces provided by this library we will introduce three types of Bonsai operator: \texttt{Estimator} objects for building and fitting models, \texttt{Predictor} objects for making predictions and \texttt{Transformer} objects for converting data.
%
Interactions between these different object types will allow a single algorithm class (implemented by \texttt{Estimator} and \texttt{Predictor} operators) to act on many different data streams.
%
An example closed-loop system allowing  control of a computer cursor with neural activity is shown represented in
Figure~\ref{fig:proposedBonsaiExtensions}b. The Bonsai
graphical program for this system is given in
Figure~\ref{fig:proposedBonsaiExtensions}c. The
objects instantiating the SpikeThresholdDetector, the KiloSortSpikeSorter and
GPFAlatentsExtractor will implement the \texttt{Transformer} interface (blue),
and the object instantiating the TimeSeriesForestClassifier will implement the
\texttt{Predictor} interface (green; Figure~\ref{fig:proposedBonsaiExtensions}c, top).
%
The parameters for the SpikeThresholdDetector and the KiloSortSpikeSorter will
be set manually by the user, as is commonly done in practice. The parameters of
the GPFAlatentsExtractor will be learnt offline in an unsupervised way by an
\texttt{Estimator} (red) in a second pipeline
(Figure~\ref{fig:proposedBonsaiExtensions}c, middle).
Finally, the parameters of the TimeSeriesForestClassifier will also be learnt
offline, but in a supervised way, by another \texttt{Estimator} in a third
pipeline (Figure~\ref{fig:proposedBonsaiExtensions}c, bottom).

Using the previous software, we will build C\# wrappers around machine
intelligence programs written in other languages, so that these programs
implement the \texttt{Estimator}, \texttt{Predictor} or \texttt{Transformer}
interfaces and can be integrated into Bonsai as machine intelligence operators.

a package management , and its community of users have
already extended Bonsai's functionality with several contributed packages.
%
However, these packages need to
be written in C\#, while most current neuroscience data analysis methods are
written in Python, R or Matlab. We propose to add to Bonsai capabilities to
communicate with software written in these languages
(Section~\ref{sec:softwareInfrastructure}).  These capabilities will allow a
large number of neuroscience data analysis methods to be easily integrated into
the Bonsai ecosystem and will provide Bonsai users a large repertoire of
advanced machine learning methods for their behavioural and neural data
analysis.
%
Using these capabilities we will integrate into Bonsai an initial set of
state-of-the art behavioural and neural data analysis methods
(Section~\ref{sec:functionality}).

% \subsubsection{Extensibility}
% \label{sec:extensibility}

% To incorporate machine learning functionality into the Bonsai ecosystem, we
% could
% implement this functionality directly into Bonsai, as done in
% \texttt{scikit-learn}. Alternatively, or in addition, we could build software
% infrastructure to facilitate the integration of existing machine learning
% software into Bonsai. We chose to begin with the second alternative because
% high-quality implementations of machine learning functionality already exist,
% which can be quickly integrated into the Bonsai ecosystem. In addition, the
% second alternative is more sustainable, as it will allow the continued addition
% of new machine learning functionality to Bonsai after the completion of the
% BBSRC-BBR funding period. Later we may decide to implement some widely-used
% machine learning functionality directly into Bonsai (e.g., for performance or
% robustness issues).

This initial set of methods should demonstrate to Bonsai users, and to
machine-learning methods developers, the potential of embedding
machine-learning functionality into the experimental control loop. We propose
user engagement activities to expose both of these groups to this
potential (Section~\ref{sec:userEngagement}).  As data-analysis method
developers recognise this opportunity, they will become interested in
contributing their methods to the Bonsai ecosystem, ensuring the
\textbf{long-term sustainability} of the proposed resource.

Adding machine learning functionality to the experimental control loop will
make possible the implementation of experiments with unprecedented levels of
control. Moreover, equipping experimental scientists without programming
experience with machine learning tools will considerably increase the
application of machine learning to the biosciences, which most probably will
translate into groundbreaking scientific discoveries.

Developers of advanced data-analysis methods should become interested in
integrating them into the Bonsai ecosystem, as it will allow their methods to
reach the wide Bonsai user community. In addition this integration will enable
method developers to easily compare the performance of their methods with
that of many methods previously integrated into the Bonsai ecosystem. Method
developers will thus become a new type of Bonsai users, which will contribute
to the \textbf{sustainability} of the proposed resource beyond the period of
BBSRC-BBR funding.

@@

\begin{figure} \begin{center}

    \href{http://www.gatsby.ucl.ac.uk/~rapela/bbsrc-bbr21/proposed_bonsai_extensions_combined.png}{\includegraphics[width=5in]{figures/proposed_bonsai_extensions_combined.png}}

    \caption{Extensions to Bonsai. (A) we propose to integrate machine learning
    functionality into the experimental control loop by building software
    infrastructure to allow Bonsai to communicate with existing machine
    learning software written in Python, R and Matlab. (B) flow chart
    representing the BCI application that we propose to implement with the
    machine learning functionality to be added to Bonsai. (C) Bonsai graphic
    program implementing the BCI application in (B).}

        \label{fig:proposedBonsaiExtensions}
    \end{center}
\end{figure}

\subsection{Software infrastructure}
\label{sec:infrastructure}

\begin{comment}
Bonsai does not currently provide machine learning functionality. For instance,
the close-loop experimental control depicted in
Figure~\ref{fig:proposedBonsaiExtensions}b cannot be currently implemented in
Bonsai. It represents the control of an experiment where a subject guides the
mouse position in a computer screen with her brain activity (Section~\ref{}).
Nodes highlighted in colour require machine intelligence functionality currently
not available in Bonsai.
%
\end{comment}

We propose to create software infrastructure to add machine intelligence
functionality to the Bonsai ecosystem in general
(Section~\ref{sec:generality}) and extensible ways
(Section~\ref{sec:extensibility}).

\begin{comment}
\begin{figure}
   \begin{center}
       \includegraphics[width=4in]{figures/bci.png}
       \caption{Close-loop system for controlling a computer cursor with brain activity}
     \label{fig:bci}
   \end{center}
\end{figure}

    The spike detector transforms membrane
potentials into spike sequences. The spike sorter transforms spikes sequences
into lists of spikes, one list per neuron. The latents extractor transforms
multi-neuron lists of spikes into latent trajectories summarising the
multi-neuron spiking activity.  The neural decoder predicts the next cursor
position from the latent trajectories.
\end{comment}

\begin{comment}
middle, whose output will set
As in this library,
there will be three type of machine intelligence operators in Bonsai:
%
estimator operators for building and fitting models, predictor operators for
making predictions and transformer operators for converting data. These
operators will be implemented by objects exposing the corresponding interfaces.

The estimator interface exposes a fit method for learning a model from training
data.  This method is called with training data (e.g., supplied as two arrays
X\_train and y\_train in supervised learning estimators). Its task is to run a
learning algorithm and to determine model-specific parameters from the training
data.  All supervised and unsupervised learning algorithms (e.g., for
classification, regression or clustering) will be offered as objects
implementing this interface. Machine learning tasks like feature extraction,
feature selection or dimensionality reduction will also be provided as
estimators.

The predictor interface extends the notion of an estimator by adding a predict
method that takes an array X\_test and produces predictions for X\_test, based
on the learned parameters of the estimator (we call the input to predict
“X\_test” in order to emphasise that predict generalises to new data). In the
case of supervised learning estimators, this method typically returns the
predicted labels or values computed by the model.

Since it is common to modify or filter data before feeding it to a learning
algorithm, the transformer interface defines a transform method. It takes as
input some new data X\_test and yields as output a transformed version of
X\_test. Preprocessing, feature selection, feature extraction and
dimensionality reduction algorithms will be provided as transformers in Bonsai.

Bonsai will provide the ability to compose new
estimators from several base estimators. Composition mechanisms can be used
to combine typical machine learning workflows into a single object which is
itself an estimator, and can be employed wherever usual estimators can be used.
Composition of estimators will be done with Pipeline objects that chain multiple estimators into a single one. This is useful
since a machine learning workflow typically involves a fixed sequence of processing steps (e.g., feature extraction, dimensionality reduction, learning and making
predictions), many of which perform some kind of learning. A sequence of N such
steps can be combined into a Pipeline if the first N-1 steps are transformers;
the last can be either a predictor, a transformer or both.

\subsection{Machine intelligence functionality to be added to Bonsai}
\label{sec:functionality}

We propose to initially add to Bonsai machine intelligence functionality to
support neuroscience experiments because, currently, neuroscience is the
research area with the largest number of Bonsai users. Within neuroscience, we
propose to add functionality for video-based behavioural analysis
(Section~\ref{sec:videoBasedBehavioralAnalysis}) and for running
brain-computer-interface (BCI) experiments (Section~\ref{sec:bci}), since these tasks
are relevant to many Bonsai users, and recent advances in machine learning
will greatly improve their implementation in Bonsai.

@@

\paragraph{Comparing multiple data-analysis methods}
\label{sec:comparisonOfMultipleMethods}

% We propose to add functionality to Bonsai to facilitate multi-method
% comparisons in user-supplied datasets.

Below we describe a comparison that we
will perform in Bonsai to asses the relative performance of the methods
described in the previous sections, using neural recordings from behaving rats.
These recordings are currently being performed, to address scientific questions,
at the laboratory of Prof.~Akrami in the SWC.
%
Details of the experimental task appear in~\citet{akramiEtAl18}.
%
Briefly, in this task there are three nose ports (left, centre, right). Rats
initiate a trial by inserting their nose in the centre port for the duration of
the fixation period, until they hear a Go stimulus. During the fixation period
two stimuli $s_a$ and $s_b$ are presented. Rats should decide which stimuli is
louder. If $s_a$ is louder than $s_b$ the correct action is to poke the nose
into the right port in order to collect a liquid reward, and if $s_b$ is louder
than $s_a$ the correct choice is the left port.

We will use high-density Neuropixel spike and LFP recordings from the rats
performing this task to learn low-dimensional representations of spiking
activity and LFPs during the decision time period (between the presentation of
the last auditory stimuli and the presentation of the Go stimulus), using the
methods described in
Sections~\ref{sec:lowDimensionalRepresentation}
and~\ref{sec:lowDimensionalRepresentationsOfLFPs}. We will also learn the
parameters of the classifiers described in Section~\ref{sec:neuralDecoding} to
optimally classify animal decisions (poke the nose into the right/left port)
based on the previous low-dimensional representations.

We will compare the decoding accuracy of all decoders. This comparison will
tells us, for the auditory working memory task, which neural data type (spikes
or LFPs), which low-dimensional representation method (e.g., LDS or Gaussian
processes), and which decoder (e.g., Bayesian or ANN) yields the best decodings
in state-of-the-art neural recordings from behaving animals.

>>>>

\begin{wrapfigure}{R}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{application/figures/extensions.pdf}
%   \input{figures/extensions}
  \caption{Proposed extensions to Bonsai}
  \label{fig:proposedBonsaiExtensions}
\end{wrapfigure}

@@

To avoid the need to re-implement a single data analysis algorithm to handle multiple different data types we will create a set of abstract machine-intelligence operators in Bonsai that can be configured to map between data substrates and machine-learning algorithms.
%
The planned abstraction layer is based on the successful \texttt{scikit-learn} Python
library \citep{buitinckEtAl13}.
%
By analogy to the interfaces provided by this library we will introduce three types of Bonsai operator: \texttt{Estimator} objects for building and fitting models, \texttt{Predictor} objects for making predictions and \texttt{Transformer} objects for converting data.
%
Interactions between these different object types will allow a single algorithm class (implemented by \texttt{Estimator} and \texttt{Predictor} operators) to act on many different data streams.
%
% An example closed-loop architecture allowing online control of a compu|ter cursor with neural activity is shown in Figure~\ref{fig:proposedBonsaiExtensions}b. The Bonsai
% graphical program for this system is given in Figure~\ref{fig:proposedBonsaiExtensions}c. The
% objects instantiating the SpikeThresholdDetector, the KiloSortSpikeSorter and
% GPFAlatentsExtractor implement the \texttt{Transformer} interface (blue),
% and the object instantiating the TimeSeriesForestClassifier implements the
% \texttt{Predictor} interface (green; Figure~\ref{fig:proposedBonsaiExtensions}c, top).
% %
% While the parameters for the SpikeThresholdDetector and the KiloSortSpikeSorter may
% be set manually by the user, the parameters of
% the GPFAlatentsExtractor can be learnt offline in an unsupervised way using an
% \texttt{Estimator} (red) in a second pipeline
% (Figure~\ref{fig:proposedBonsaiExtensions}c, middle).
% Finally, the parameters of the TimeSeriesForestClassifier can also be learnt
% offline, but in a supervised way, using another \texttt{Estimator} in a third
% pipeline (Figure~\ref{fig:proposedBonsaiExtensions}c, bottom).

% This initial set of methods should demonstrate to Bonsai users, and to
% machine-learning methods developers, the potential of embedding
% machine-learning functionality into the experimental control loop. We propose
% user engagement activities to expose both of these groups to this
% potential (Section~\ref{sec:userEngagement}).  As data-analysis method
% developers recognise this opportunity, they will become interested in
% contributing their methods to the Bonsai ecosystem, ensuring the
% \textbf{long-term sustainability} of the proposed resource.

% Adding machine learning functionality to the experimental control loop will
% make possible the implementation of experiments with unprecedented levels of
% control. Moreover, equipping experimental scientists without programming
% experience with machine learning tools will considerably increase the
% application of machine learning to the biosciences, which most probably will
% translate into groundbreaking scientific discoveries.

% Developers of advanced data-analysis methods should become interested in
% integrating them into the Bonsai ecosystem, as it will allow their methods to
% reach the wide Bonsai user community. In addition this integration will enable
% method developers to easily compare the performance of their methods with
% that of many methods previously integrated into the Bonsai ecosystem. Method
% developers will thus become a new type of Bonsai users, which will contribute
% to the \textbf{sustainability} of the proposed resource beyond the period of
% BBSRC-BBR funding.

@@

% \begin{figure} \begin{center}

%     \href{http://www.gatsby.ucl.ac.uk/~rapela/bbsrc-bbr21/proposed_bonsai_extensions_combined.png}{\includegraphics[width=5in]{figures/proposed_bonsai_extensions_combined.png}}

%     \caption{Extensions to Bonsai. (A) we propose to integrate machine learning
%     functionality into the experimental control loop by building software
%     infrastructure to allow Bonsai to communicate with existing machine
%     learning software written in Python, R and Matlab. (B) flow chart
%     representing the BCI application that we propose to implement with the
%     machine learning functionality to be added to Bonsai. (C) Bonsai graphic
%     program implementing the BCI application in (B).}

%         
%     \end{center}
% \end{figure}

\subsection{Machine intelligence functionality}
\label{sec:functionality}

We will use the frameworks developed above to integrate existing machine-intelligence programs within one or more Bonsai packages.  Based on user input, we have identified initial domains of video-based behavioural analysis (Section~\ref{sec:videoBasedBehavioralAnalysis}) and
brain-computer-interface (BCI) experiments (Section~\ref{sec:bci}).

@@

\subsection{Comparing multiple data-analysis methods}
\label{sec:comparisonOfMultipleMethods}

We propose to add functionality to Bonsai to facilitate multi-method
comparisons in user-supplied datasets.

% Below we describe a comparison that we
% will perform in Bonsai to asses the relative performance of the methods
% described in the previous sections, using neural recordings from behaving rats.
% These recordings are currently being performed, to address scientific questions,
% at the laboratory of Prof.~Akrami in the SWC.
% %
% Details of the experimental task appear in~\citet{akramiEtAl18}.
% %
% Briefly, in this task there are three nose ports (left, centre, right). Rats
% initiate a trial by inserting their nose in the centre port for the duration of
% the fixation period, until they hear a Go stimulus. During the fixation period
% two stimuli $s_a$ and $s_b$ are presented. Rats should decide which stimuli is
% louder. If $s_a$ is louder than $s_b$ the correct action is to poke the nose
% into the right port in order to collect a liquid reward, and if $s_b$ is louder
% than $s_a$ the correct choice is the left port.

% We will use high-density Neuropixel spike and LFP recordings from the rats
% performing this task to learn low-dimensional representations of spiking
% activity and LFPs during the decision time period (between the presentation of
% the last auditory stimuli and the presentation of the Go stimulus), using the
% methods described in
% Sections~\ref{sec:lowDimensionalRepresentation}
% and~\ref{sec:lowDimensionalRepresentationsOfLFPs}. We will also learn the
% parameters of the classifiers described in Section~\ref{sec:neuralDecoding} to
% optimally classify animal decisions (poke the nose into the right/left port)
% based on the previous low-dimensional representations.

% We will compare the decoding accuracy of all decoders. This comparison will
% tells us, for the auditory working memory task, which neural data type (spikes
% or LFPs), which low-dimensional representation method (e.g., LDS or Gaussian
% processes), and which decoder (e.g., Bayesian or ANN) yields the best decodings
% in state-of-the-art neural recordings from behaving animals.

### 2021-11-17 application/resourceManagementandSustainability.tex
Updates from Goncalo

\section{Resource management}
\textcolor{red}{I'm making this up -- Goncalo please comment/correct.}
%
As an open source environment developed over almost 10 years, Bonsai 
has well-established user-community-based management structures.
%
The core system and the majority of extension packages are hosted on a
publicly accessible website at \url{bonsai-rx.org}.
%
Package installation and updates are supported by standalone package
management software included within the core Bonsai distribution.
%
This ensures that bugfixes and enhancements are easily available but
under full user control (to ensure working experiments are not
disrupted by changes in the system).

Best-practice code review procedures are followed prior to including
package updates.  \textcolor{red}{Goncalo/Joaquin -- could you expand
  on this?}

@@

The software engineers recruited to this project will work with our
collaborators at Neurogears Ltd, to strengthen and maintain these
practices.  Once fully familiar with the Bonsai core architecture they
will participate in review of pull requests and other code
contributions.

@@

Bonsai is a freely available open-source resource, already widely adopted within the systems neuroscience and behavioural science communities (see Fig~\ref{fig:bonsai}).  
%
Current contributors include other large open projects (OpenEphys, \textcolor{red}{\dots}), as well as companies developing and supplying new hardware devices  (\textcolor{red}{\dots}).
%
This level of community uptake, engagement, and commitment provides a foundation for long-term sustainability of the resource.

Neurogears Ltd is a keystone of this community and a key a collaborator in the current proposal.  As a company founded by the original developer of Bonsai, it helps to support and expand the user base well beyond the life of the current proposal.  Neurogears provides quality control and curation for core Bonsai code.  \textcolor{red}{other ways NG contributes ...}

The specific contributions developed as part of the current project will form new components within the core Bonsai package. 
%
They will be developed with input from the advisory panel, early public distribution, user testing within the SWC, and planned workshops, ensuring close user engagement from the beginning of the development cycle and so building momentum for their use, support and dissemination.
%
\textcolor{red}{Can we say that NG will take over responsibility for maintenance and support beyond the project?}

>>>>

\section{Resource management}
\textcolor{red}{I'm making this up -- Goncalo please comment/correct.}
%
As an open source project developed over almost 10 years, Bonsai 
has well-established management practices built around the user community.
%
The source code for the core system and the majority of extension packages
is hosted on GitHub at \url{github.com/bonsai-rx}.
%
The documentation, including learning materials and all relevant community
links are published in a publicly accessible website at \url{bonsai-rx.org}.
%
A community user forum is available as a public Google Group with an active
user base of more than 700 users (Figure~\ref{fig:bonsai}f) and several
new questions and answers exchanged every day.
%
Package installation and updates are supported by standalone package
management software included within the core Bonsai distribution and hosted
in a shared centralized and curated repository.
%
This ensures that bugfixes and enhancements are made easily available but
also importantly allows for reproducible deployment of user environments
to ensure working experiments are not disrupted by changes in the system.
Large-scale experimental projects such as the International Brain Laboratory
have leveraged this capability of Bonsai to standardize data acquisition and
control software across dozens of laboratories around the world.
%
Community technology previews are made available through the package manager,
providing early adopters the possibility of opting-in to the latest features
prior to making them generally available. This allows for the community
to help identify bugs and report issues with new features on GitHub, thereby
distributing bug reporting and triage and overall improving the quality of
stable releases.

@@

The core compiler and IDE is extensively tested, and also takes advantage of
a strong type system to greatly reduce bugs at compile-time. A set of best
practices for leveraging this infrastructure for building new packages is
in place.

The software engineers recruited to this project will work with our
collaborators at NeuroGEARS Ltd, to strengthen and maintain these
practices.  Once fully familiar with the Bonsai core architecture they
will participate in review of pull requests and other code contributions.

@@

Bonsai is a freely available open-source resource, already widely adopted within
the systems neuroscience and behavioural science communities (see Fig~\ref{fig:bonsai}).  
%
Current contributors include other large open projects (OpenEphys, DeepLabCut),
as well as companies developing and supplying new hardware devices  (Neurophotometrics).
%
This level of community uptake, engagement, and commitment provides a foundation for
long-term sustainability of the resource.

NeuroGEARS Ltd is a keystone of this community and a key collaborator in the current proposal.
As a company founded by the original developer of Bonsai, it helps to support and expand the
user base well beyond the life of the current proposal. NeuroGEARS currently provides
quality control and curation for the core Bonsai code, as well as providing training and support
services for research institutions around the world who want to start using Bonsai.
Finally, NeuroGEARS also facilitates knowledge transfer across users and institutions adopting
Bonsai, by moderating the community user forums and collaborating with organisations
who want to build new innovative research projects on top of Bonsai, as is the case with the
current proposal.

The specific contributions developed as part of the current project will form new core
components and packages within the Bonsai ecosystem. The new language integration packages are
of particular relevance to the long-term sustainability of Bonsai, as they will dramatically
increase the interoperability of Bonsai with existing software development communities which are
already well established in basic science (MATLAB, Python, R).
%
The opportunity to develop these new packages with input from the core advisory panel of Bonsai,
and early public distribution and user testing within the SWC and planned workshops, will ensure
solid and broad user engagement from the beginning of the development cycle and will therefore
build momentum for their use, support and dissemination.
%
The close collaboration with the core maintainers of the Bonsai ecosystem at NeuroGEARS will
allow them to take over responsibility for maintenance and support of these packages beyond the
end of the project, following the same principles of open-source and free sharing of knowledge
that characterises the entire Bonsai ecosystem.
